---
title: "Multivariate And Categorical Data Analysis - 2"
author: "Ritesh.kumar2"
date: "9/20/2018"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r claenup workspace, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
gc()
```

```{r load Library, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
library(MASS)
library(mixtools)

#alibrary(gRain)
#source("https://bioconductor.org/biocLite.R")
#biocLite("RBGL")
library(RBGL)
library(gRbase)
library(gRain)
library(Rgraphviz)
#biocLite("Rgraphviz")
library(igraph)
library(ggm)
```

# Assignment - 2

Q1) [Marks 4+4+4+5+4+2+2+3=28]
For this question you will be using the “AIMSNingalooReefAirPressure.csv” dataset.
This dataset gives the air pressure measurements collected at Ningaloo reef in Western Australia over a one year period between August 2017 and August 2018.
You can download this dataset from the Assignment folder in CloudDeakin. You can use the following R code to load the data:
AIMSDataAirPres<- as.matrix(read.csv("AIMSNingalooReefAirPressure.csv", header = TRUE, sep = ",", quote = "\"", dec = ".", fill = TRUE, comment.char = ""))

### 1.1) Provide a time series plot of the data (use the index as the time (x-axis)). Use the following R code to plot it:
  plot(AIMSDataAirPres)
  Provide the five point summary, mean and the standard deviation of the air pressure data.

```{r Q1.1}

AIMSDataAirPres <- 
  as.matrix(read.csv(
    "data/AIMSNingalooReefAirPressure.csv", 
    header = TRUE, 
    sep = ",", 
    quote = "\"", 
    dec = ".", fill = TRUE, comment.char = ""))

plot(AIMSDataAirPres)

message(sprintf(
  "\nFive Number Summary for 'Air Pressure Measurements': 
  \tMinimum value  = %s
  \tLower Quartile = %s
  \tMedian = %s
  \tHiger Quartile = %s
  \tMaximum Value = %s
  Mean value  = %s
  Standard Deviation = %s", 
  min(AIMSDataAirPres), 
  quantile(AIMSDataAirPres, 0.25),
  median(AIMSDataAirPres),
  quantile(AIMSDataAirPres, 0.75),
  max(AIMSDataAirPres),
  mean(AIMSDataAirPres), 
  sd(AIMSDataAirPres, 0.25)))


#View(AIMSDataAirPres.df)
```

### 1.2) Plot the histogram of the air pressure data. Comment on the shape. How many modes can be observed in the data?

```{r Q1.2}

AIMSDataAirPres <- 
  as.matrix(read.csv(
    "data/AIMSNingalooReefAirPressure.csv", 
    header = TRUE, 
    sep = ",", 
    quote = "\"", 
    dec = ".", 
    fill = TRUE, 
    comment.char = ""))

AIMSDataAirPres.df <- as.data.frame(AIMSDataAirPres)
colnames(AIMSDataAirPres.df) <- c("value")

ggplot(
  data = AIMSDataAirPres.df,
  aes(AIMSDataAirPres.df$value)) +
    geom_histogram(
      breaks = seq(995, 1025, by = 1),
      #Bandwidth = 1.407,
      col = "red",
      fill = "green",
      alpha = .2) +
    labs(title = "Histogram for Air Pressure Measurements") +
    labs(x = "Air Pressure Measurements", y="Frequency") + 
    xlim(c(995,1025)) + 
    ylim(c(0,40))

```

#### Comments:
- The histogram is slightly left skewed
- It is not having any outlierers 
- There are two modes


### 1.3) Fit a single Gaussian model to the distribution of the data, where is the mean and  is the standard deviation of the Gaussian distribution
Find the maximum likelihood estimate (MLE) of the parameters, i.e., the mean and the standard deviation. You can use the following code to perform the fitting:
  fit1 <- fitdistr(AIMSDataAirPres,"normal")
  Plot the obtained density distribution.

```{r Q1.3}

fit1 <- fitdistr(AIMSDataAirPres,"normal")

message(sprintf(
  "\n The maximum likelihood estimate(MLE): \n\tMean = %s\n\tStandard Deviation = %s", 
  fit1$estimate[1], fit1$estimate[2])) 


plot(density(AIMSDataAirPres))

```

### 1.4) Fit a mixture of Gaussians model to the distribution of the data using the number of Gaussians equal to the number of modes found in the data (in Q1.2 above).  Write the R code to perform this. Provide the mixing coefficients, mean and standard deviation for each of the Gaussians found.

```{r Q1.4}

mixmdl <- normalmixEM(AIMSDataAirPres, k=2)
#mixmdl
summary(mixmdl)

message(sprintf(
  "Mixing coefficients for 1st Gaussian:\n\tMean value  = %s\n\tStandard Deviation = %s\nMixing coefficients for 2nd Gaussian:\n\tMean value  = %s\n\tStandard Deviation = %s", 
  mixmdl$mu[1], 
  mixmdl$sigma[1],
  mixmdl$mu[2], 
  mixmdl$sigma[2]))

```

### 1.5) Plot these Gaussians on top of the histogram plot. Include a plot of the combined density distribution as well (use different colors for the density plots in the same graph).

```{r Q1.5}

plot(mixmdl,which=2)
lines(density(AIMSDataAirPres), lty=2, lwd=2)
mixmdl$lambda
mixmdl$mu
mixmdl$sigma

```

### 1.6) Provide a plot of the log likelihood values obtained over the iterations and comment on them.

```{r Q1.6}

plot(mixmdl$all.loglik)

```

### 1.7) Comment on the distribution models obtained in Q1.3 and Q1.4. Which one is better?

```{r Q1.7}


```

### 1.8) What is the main problem that you might come across when performing a maximum likelihood estimation using mixture of Gaussians? How can you resolve that problem in practice?

```{r Q1.8}


```

## Q3) [Marks 4+5+3 = 12] A belief network models the relation between the variables oil; inf; eh; bp; rt which stand for the price of oil, inflation rate, economy health, British Petroleum Stock price, and retailer stock price. Each variable takes the states low; high, except for bp and rt which have states low; high; normal. The belief network model for these variables has tables as shown below:

### 3.2) Use the below libraries in R to create this belief network in R along with the probability values as shown in the above table. 
  You may use the following libraries for this: 
  
  source("https://bioconductor.org/biocLite.R")
  
  biocLite("RBGL")
  
  library(RBGL)
  
  library(gRbase)
  
  library(gRain)
  
  library(Rgraphviz)

  define the appropriate network and use the “compileCPT()”function to Compile list of conditional probability tables and create the network.
  Show the probability tables obtained in the R output and verify with the above table.

```{r Q3.2}
lh <- c("low", "high")
lhn <- c("low", "high", "normal")

eh <- cptable(~eh, values = c(20,80), levels=lh)
bp <- cptable(~bp|oil, values = c(80,5,15, 10,50,40), levels=lhn)
oil <- cptable(~oil|eh, values = c(90,10,5,95), levels = lh)
rt.inf_eh <- cptable(~rt|inf:eh, values = c(60,10,30,20,60,20,10,70,20,5,85,10), levels = lhn)
inf.oil_eh <- cptable(~inf|oil:eh, values = c(90,10,20,80,10,90,2,98), levels = lh)

#Compile list of conditional probability tables and create the network:
plist <- compileCPT(list(eh,bp,oil,rt.inf_eh,inf.oil_eh))
plist$rt

net1 <- grain(plist)
net1
summary(net1)

plot(net1)

```

### 3.3) Use R program to compute the following:
Given that the BP stock price is high and the retailer stock price is normal, what is the probability that inflation is high?

```{r Q3.3}

net12 <- setEvidence(net1, evidence=list(bp="high", rt="normal"))
result <- querygrain( net12, nodes=c("inf") )
message(sprintf(
  "p(inf = high | BP stock price = high, retailer stock price = normal) = %s", 
  result$inf[2]))
```

## Q4) [Marks : (2+2+2+2+2+3+3) + 5=21]. Consider the Bayesian network shown in the diagram below.

### 4.2) Write a R-Program to produce this Bayesian network and perform the d-sepration tests for all of the above cases mentioned in Q4.1 (a) to (g).

```{r Q4.2}

d.sep <- DAG(D ~ A+B, F ~ D+C+E, H ~ E+F, C ~ A, G ~ C)
drawGraph(d.sep, adjust = FALSE)

message(sprintf(
  "C is marginally independent of G = %s
C is conditionally independent of H given E = %s
G is conditionally independent of E given D = %s
C is conditionally independent of H given F = %s
B is conditionally independent of G given F = %s
B is conditionally independent of G given D, C, E = %s
A is conditionally independent of H given D, F = %s", 
  dSep(d.sep, first="C", second="G", cond = c()),
  dSep(d.sep, first="C", second="H", cond = c("E")),
dSep(d.sep, first="G", second="E", cond = c("D")),
dSep(d.sep, first="C", second="H", cond = c("F")),
dSep(d.sep, first="B", second="G", cond = c("F")),
dSep(d.sep, first="B", second="G", cond = c("D","C","E")),
dSep(d.sep, first="A", second="H", cond = c("D","F"))))
  
```