---
title: "Multivariate And Categorical Data Analysis"
author: "Ritesh Kumar (218211529)"
date: "8/8/2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r claenup workspace, message=FALSE, warning=FALSE, include=FALSE}
rm(list=ls())
gc()
```

```{r load Library, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
```

# Assignment - 1
Bike sharing dataset (BikeShare)
This dataset gives the count of bikes rented between 11am - 12pm on different days and locations through the Capital Bikeshare System (operating in US cities) between 2011 and 2012. The variables include the following (9 variables):

Season: Categorical: 1 = Spring, 2 = Summer, 3 = Autumn (fall), 4 = Winter

Working day: 0 = Weekend, 1 = Workday.

Weather: Categorical variable

  1: Clear, Few clouds, Partly cloudy, Partly cloudy
  
  2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
  
  3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered cloud 
  
  4: Heavy Rain + Ice Pallets + Thunderstorm Mist, Snow + Fog
  
Temperature: Temperature in Celsius.

`Feeling' Temperature: `Feels like' temperature, reported in Celsius.

Humidity: Humidity (given as a percentage).

Windspeed: Windspeed (measured in km/h).

Casual users: Count of casual users that used a bike at that time.

Registered users: Count of registered users that used a bike at that time.

### Question-1
Download the txt file “BikeShareTabSep.txt” and save it to your R working directory.

• Assign the data to a matrix, e.g. using the.data<-as.matrix(read.table("BikeShareTabSep.txt"))

• Generate a sample of 400 data using the following: my.data <- the.data [sample(1:727,400),c(1:9)]

Save “my.data” to a text file titled “name-StudentID-BikeShareMyData.txt" using the following R code (NOTE: you must upload this text file with your submission).

write.table(my.data,"name-StudentID-BikeShareMyData.txt")
Use the sampled data (“my.data”) to answer the following questions.

```{r Q1}
the.data<-as.matrix(read.table("data/BikeShareTabSep.txt"))
#nrow(the.data)
my.data <- the.data [sample(1:727,400),c(1:9)]
write.table(my.data,"data/ritesh-218211529-BikeShareMyData.txt")

```

#### 1.1) Draw histograms for ‘Registered users’ and ‘Temperature’ values, and comment on them. [3 Marks]

```{r Q1.1}
# Column Names
col.names <- 
  c("Season", "WorkingDay", "Weather", "Temperature", "FeelingTemperature", 
    "Humidity", "Windspeed", "CasualUsers", "RegisteredUsers")
# Set the column names
colnames(my.data) <- col.names
my.data.df <- as.data.frame(my.data)

# Convert columns to categorical type
my.data.df$Season <-as.factor(my.data.df$Season)
my.data.df$WorkingDay <-as.factor(my.data.df$WorkingDay)
my.data.df$Weather <-as.factor(my.data.df$Weather)

# Data Summary
summary(my.data.df)
#View(my.data.df)
```

```{r Registered User Histogram}
# Registered User Histogram
ggplot(
  data = my.data.df,
  aes(my.data.df$RegisteredUsers)) +
    geom_histogram(
      breaks = seq(0, 500, by = 20),
      col = "red",
      fill = "green",
      alpha = .2) +
    labs(title = "Histogram for Registered Users") +
    labs(x = "Registered Users", y="Count") + 
    xlim(c(0,500)) + 
    ylim(c(0,60))
```

```{r Temperature Histogram}
# Temperature Histogram
ggplot(
  data = my.data.df,
  aes(my.data.df$Temperature)) +
    geom_histogram(
      breaks = seq(0, 50, by = 2),
      col = "red",
      fill = "green",
      alpha = .2) +
    labs(title = "Histogram for Temperatures") +
    labs(x = "Temperature", y="Count") + 
    xlim(c(0,50)) + 
    ylim(c(0,50))
```

####  1.2) Give the five number summary and the mean value for the ‘Casual users’ and the ‘Registered users’ separately. [3 Marks]

```{r Q1.2}
message("Five Number Summary for 'Casual Users': ")
message(sprintf(
  "Minimum value  = %s\nLower Quartile = %s\nMedian = %s\nHiger Quartile = %s\nMaximum Value = %s\n", 
  min(my.data.df$CasualUsers), 
  quantile(my.data.df$CasualUsers, 0.25),
  median(my.data.df$CasualUsers),
  quantile(my.data.df$CasualUsers, 0.75),
  max(my.data.df$CasualUsers)))

message("Mean for 'Casual Users': = ", mean(my.data.df$CasualUsers))

message("\nFive Number Summary for 'Registered users': ")
message(sprintf(
  "Minimum value  = %s\nLower Quartile = %s\nMedian = %s\nHiger Quartile = %s\nMaximum Value = %s\n", 
  min(my.data.df$RegisteredUsers), 
  quantile(my.data.df$RegisteredUsers, 0.25),
  median(my.data.df$RegisteredUsers),
  quantile(my.data.df$RegisteredUsers, 0.75),
  max(my.data.df$RegisteredUsers)))

message("Mean for 'Registered users' = ", mean(my.data.df$RegisteredUsers))

```

####  1.3) Draw a parallel Box plot using the two variables; ‘Casual users’ and the ‘Registered users’. Use the answers to Q1.2 and the Boxplots to compare and comment on them. [3 Marks]

```{r Q1.3}

boxplot(
  my.data.df[c("CasualUsers", "RegisteredUsers")], 
  main="Distribution of Users",
  xlab="Types of User", 
  ylab="User Count",
  col = c("red","sienna"),
  at = c(1,2), 
  #par(mar = c(12, 5, 4, 2) + 0.1),
  names = c("CasualUsers", "RegisteredUsers"))


```

####  1.4) Draw a scatterplot of ‘Temperature’ and ‘Casual users’ for the first 200 data vectors selected from the “my.data” (name the axes) and comment on them [2 Marks]

```{r Q1.4}
ggplot(head(my.data.df,200), aes(x=Temperature, y=CasualUsers)) + 
  geom_point()
  
#scatter.smooth(x=head(my.data.df,200)$Temperature, y=head(my.data.df,200)$CasualUsers, main="CasualUsers ~ Temperature")
```


### 1.5) Fit a linear regression model to the ‘temperature’ (as x) and the ‘casual users’ (as y) using the first 200 data vectors selected from the “my.data”. Write down the linear regression equation. Plot the line on the same scatter plot. Compute the correlation coefficient and the coefficient of Determination. Explain what these results reveal. [5 Marks]

```{r Q1.5}
cor(my.data.df$Temperature, my.data.df$CasualUsers) 

linearMod <- lm(Temperature ~ CasualUsers, data=my.data.df)
summary(linearMod)

{plot(CasualUsers ~ Temperature, data=head(my.data.df,200)) 
abline(lm(CasualUsers ~ Temperature, data=head(my.data.df,200)))}

```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
